Associahedra for finite type cluster algebras \\ and minimal relations between $\b{g}$-vectors

### 2019-04-09 thirdPaper/draft/typeConeAssociahedra.tex
intro CA

% formating the part command
\makeatletter
\def\part{\@startsection{part}{1}%
\z@{.7\linespacing\@plus\linespacing}{.8\linespacing}%
{\LARGE\sffamily\centering}}
\@addtoreset{section}{part}
\makeatother
\renewcommand{\thesection}{\arabic{part}.\arabic{section}}

@@

@@

@@

The associahedron appears as a fundamental structure in several mathematical theories, such as moduli spaces and topology~\cite{Stasheff, Keller-AinfinityAlgebras}, operads and rewriting theory~\cite{}, combinatorial Hopf algebras~\cite{LodayRonco, ChatelPilaud, Pilaud-brickAlgebra}, diagonal harmonics~\cite{}, mathematical physics~\cite{ArkaniHamedBaiHeYan}, etc.
Another striking illustration of the ubiquity of the associahedron lies in the theory of cluster algebras introduced by S.~Fomin and A.~Zelevinsky in~\cite{FominZelevinsky-ClusterAlgebrasI} with motivation coming from total positivity and canonical bases.
Cluster algebras are commutative rings generated by variables obtained from an initial seed by the discrete dynamical process of seed mutations.
Finite type cluster algebras, whose mutation graph is finite, were classified in~\cite{FominZelevinsky-ClusterAlgebrasII} using the Cartan-Killing classification for crystallographic root systems.
Finite type cluster complexes were geometrically realized first by F.~Chapoton, S.~Fomin and A.~Zelevinsky~\cite{ChapotonFominZelevinsky} based on the $\b{d}$-vector fans~\cite{FominZelevinsky-YSystems, FominZelevinsky-ClusterAlgebrasII}, then by~C.~Hohlweg, C.~Lange and H.~Thomas~\cite{HohlwegLangeThomas} based on the Cambrian fans~\cite{Reading-CambrianLattices, ReadingSpeyer}, and finally by C.~Hohlweg, V.~Pilaud and S.~Stella~\cite{HohlwegPilaudStella} based on the $\b{g}$-vector fans~\cite{FominZelevinsky-ClusterAlgebrasIV} with respect to an arbitrary initial seed.
The resulting polytopes are known as generalized associahedra, and the classical associahedra corresponds to the cluster algebras of type~$A$.

This paper focuses on a recent construction of the associahedron that appeared in mathematical physics.
Motivated by the prediction of the behavior of scattering particules, N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan recently described in~\cite[Sect.~3.2]{ArkaniHamedBaiHeYan} the kinetic associahedron.
It is a class of polytopal realizations of the classical associahedron obtained as sections of a high-dimensional positive orthant with well-chosen affine subspaces.
This construction provides a large degree of freedom in the choice of the parameters defining these affine subspaces and produces all polytopes whose normal fan is affinely equivalent to that of J.-L.~Loday's associahedron~\cite{Loday} (see \cref{subsec:typeConeAsso}).
These realizations were then extended by V.~Bazier-Matte, G.~Douville, K.~Mousavand, H.~Thomas and E.~Yildirim~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the context of finite type cluster algebras using tools from representation theory of quivers.
More precisely, they fix a finite type cluster algebra~$\c{A}$ and consider the real euclidean space~$\R^\c{V}$ indexed by the set~$\c{V}$ of cluster variables of~$\c{A}$.
Starting from an acyclic initial seed~$\seed_\circ$, they consider a certain set~$\c{M}$ of mesh relations and compute the corresponding linear relations between the $\b{g}$-vectors, interpreted as linear spaces in~$\R^\c{V}$.
Finally, they perturb these mesh linear spaces by a collection~$\b{\ell} \in \R^{\c{M}}$ of positive constants and intersect the resulting perturbed affine mesh relations with the positive orthant of~$\R^\c{V}$ (see \cref{subsec:typeConeCA}).
They show that the resulting polytope is always a generalized associahedron, and that its normal fan is the $\b{g}$-vector fan of~$\c{A}$ with respect to the initial acyclic seed~$\seed_\circ$.
A by-product of their construction is that the space of all polytopal realizations of the $\b{g}$-vector fan of~$\c{A}$ with respect to the initial acyclic seed~$\seed_\circ$ is a simplicial cone.

In this paper, we revisit, extend and explore further this construction using a different approach.
Given a complete simplicial fan~$\Fan$, we consider its space~$\typeCone(\Fan)$ of polytopal realizations, called type cone of~$\Fan$ in~\cite{McMullen-typeCone}.
This type cone is known to be a polyhedral cone defined by a collection of inequalities corresponding to the linear dependences among the rays of~$\Fan$ contained in pairs of adjacent maximal cones of~$\Fan$.
Our approach is based on a simple but crucial observation: if the type cone~$\typeCone(\Fan)$ is a simplicial cone, then all polytopal realizations of~$\Fan$ can be described as sections of a high dimensional positive orthant with a collection of affine subspaces corresponding to the facets of~$\typeCone(\Fan)$ (see \cref{coro:simplicialTypeCone}).
To prove that the type cone~$\typeCone(\Fan)$ is simplicial, we just need to identify which pairs of adjacent maximal cones of~$\Fan$ correspond to the facets of~$\typeCone(\Fan)$ and to show that the corresponding linear dependences among their rays positively span the linear dependence among the rays of any pair of adjacent maximal cones of~$\Fan$.

%
%
%Their proof, based on representation theory of quivers, is subtle: on the one hand, only simply-laced types are explicitly treated and the remaining finite type are argued by folding; on the other hand, the technique requires to work with rational perturbations~$\b{\ell}$ and to conclude for arbitrary real perturbations by an approximation argument.

@@

@@

\part{Type cones of $\b{g}$-vector fans}

@@

\begin{definition}\label{def:typecone}
The \defn{type cone} of~$\Fan$ is the cone~$\typeCone(\Fan)$ of all $\Fan$-admissible height vectors~$\b{h}$:
\begin{align*}
\typeCone(\Fan) & \eqdef \set{\b{h} \in \R^N}{\Fan \text{ is the normal fan of } P_\b{h}} \\
& = \Bigset{\b{h} \in \R^N}{\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0 \; \begin{array}{l} \text{for any adjacent maximal} \\ \text{cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$} \end{array}}.
\end{align*}
\end{definition}

@@

Although we will only treat with simplicial fans, for completeness, note that \cref{prop:characterizationPolytopalFan} and \cref{def:typecone} can be easily adapted to non-simplicial fans. To describe the type cone of an arbitrary complete fan, it suffices to consider any simplicial refinement, and to set some of the strict inequalities of the definition of the type cone to equality.
\begin{proposition}
 \label{prop:non-simplicial}
 Let $\Fan'$ be a complete fan that coarsens the essential complete simplicial fan~$\Fan$.
 The \defn{type cone}~$\typeCone(\Fan')$ of all $\Fan'$-admissible height vectors~$\b{h}$ is
 \begin{align*}
 \typeCone(\Fan) 
%  & \eqdef \set{\b{h} \in \R^N}{\Fan \text{ is the normal fan of } P_\b{h}} \\
& = \left\{\b{h} \in \R^N\;\middle|\;
\begin{array}{ll}
 \sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} = 0
 & \begin{array}{l} \text{for any adjacent maximal cones} \\ \text{$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ \textbf{belonging}}\\\text{to the same maximal cone of $\Fan'$; and}\end{array}
\\\\
\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0
 & \begin{array}{l} \text{for any adjacent maximal cones} \\ \text{$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ \textbf{not belonging}}\\\text{to the same maximal cone of $\Fan'$}\end{array}
\end{array}
 \right\}.
 \end{align*}
\end{proposition}
\arnau{should add proof or citation}

@@

\begin{example}\label{ex:typecone}
Consider the $2$-dimensional fan~$\Fan$ depicted in \cref{fig:typecone} (left) (we will encounter this fan again in \cref{exm:typeConeAsso}). It has five rays $v_1,\dots,v_5$ and five maximal cones $a,\dots, e$. Its type cone lies in~$\R^5$, but has a $2$-dimensional linearity space. It suffices to study $\b{K}\ctypeCone(\Fan)$ which is an essential simplicial cone in~$\R^3$. One can further reduce the dimension by intersecting with a transversal hyperplane, to get a $2$-dimensional representation. This is the triangle that is depicted in \cref{fig:typecone} (right). The five pairs of adjacent maximal cones give rise to five defining inequalities for $\ctypeCone(\Fan)$, although only three are non-redundant. If $H_{ij}$ is the hyperplane supporting the inequality given by the two adjacent maximal cones~$i$ and~$j$, then only $H_{ab}$, $H_{cd}$ and $H_{bc}$ support facets of the type cone. The relative interior of the three facets of~$\ctypeCone(\Fan)$ are the type cones of three coarsenings of~$\Fan$, obtained by merging certain adjacent maximal cones (those that are extremal). One of the rays generating~$\b{K}\ctypeCone(\Fan)$ corresponds to a $2$-dimensional simplex, and its normal fan is obtained by merging two pairs of extremal adjacent maximal cones. The two remaining rays lie not only in the hyperplanes~$H_{ij}$ defining the facets it belongs to, but also one extra such hyperplane. Its normal fan has only two cones, one arising as a merging of three cones and one arising as a merging of two cones. These fans are not essential, they have a one-dimensional linearity. Hence, the corresponding polytopes are not full-dimensional, they are of codimension~$1$.

\begin{figure}[htpb]
	\capstart
	\centerline{{\includegraphics[width=.3\linewidth]{typeconeFan}}\hspace{1.5cm}\includegraphics[width=.4\linewidth]{typecone}}
	\caption{A complete $2$-dimensional fan~$\Fan$ with five rays $v_1,\dots,v_5$ and five maximal cones $a,\dots, e$. We depict the intersection of $\b{K}\ctypeCone(F)$ with a hyperplane to get a $2$-dimensional slice. On top of each face of $\ctypeCone(F)$, and example of polytope with the corresponding normal fan is depicted.}
	\label{fig:typecone}
\end{figure}

@@

\begin{figure}[h]
    \begin{adjustbox}{center}
        \begin{tikzpicture}
        	\matrix (m) [matrix of math nodes, row sep=.35cm, column sep=.2cm, nodes={anchor=center, align=center, inner sep=0pt}]{
        		& \;\includegraphics[scale=.7]{diagonalA2} & \node (b) {\red \circled{B}}; & \;\includegraphics[scale=.7]{diagonalA4} & \\
        		\;\includegraphics[scale=.7]{diagonalA1} & \node (a) {\red \circled{A}}; & \;\includegraphics[scale=.7]{diagonalA3} & \node (c) {\red \circled{C}}; & \;\includegraphics[scale=.7]{diagonalA5} \\};
        	\draw[->] (m-2-1) -- (m-1-2);
        	\draw[->] (m-1-2) -- (m-2-3);
        	\draw[->] (m-2-3) -- (m-1-4);
        	\draw[->] (m-1-4) -- (m-2-5);
        	\draw[red] (a) -- (m-2-1);
        	\draw[red] (a) -- (m-2-3);
        	\draw[red] (a) -- (m-1-2);
        	\draw[red] (b) -- (m-1-2);
        	\draw[red] (b) -- (m-1-4);
        	\draw[red] (b) -- (m-2-3);
        	\draw[red] (c) -- (m-2-3);
        	\draw[red] (c) -- (m-2-5);
        	\draw[red] (c) -- (m-1-4);
        \end{tikzpicture}
        \quad
        \begin{tikzpicture}
        	\matrix (m) [matrix of math nodes, row sep=.35cm, column sep=.2cm, nodes={anchor=center, align=center, inner sep=0pt}]{
        		&& \;\includegraphics[scale=.7]{diagonalB3} & \node (c) {\red \circled{C}}; & \;\includegraphics[scale=.7]{diagonalB7} && \\
        		& \;\includegraphics[scale=.7]{diagonalB2} & \node (b) {\red \circled{B}}; & \;\includegraphics[scale=.7]{diagonalB5} & \node (e) {\red \circled{E}}; & \;\includegraphics[scale=.7]{diagonalB8} & \\
        		\;\includegraphics[scale=.7]{diagonalB1} & \node (a) {\red \circled{A}}; & \;\includegraphics[scale=.7]{diagonalB4} & \node (d) {\red \circled{D}}; & \;\includegraphics[scale=.7]{diagonalB6} & \node (f) {\red \circled{F}}; & \;\includegraphics[scale=.7]{diagonalB9} \\};
        	\draw[->] (m-3-1) -- (m-2-2);
        	\draw[->] (m-2-2) -- (m-1-3);
        	\draw[->] (m-2-2) -- (m-3-3);
        	\draw[->] (m-1-3) -- (m-2-4);
        	\draw[->] (m-3-3) -- (m-2-4);
        	\draw[->] (m-2-4) -- (m-1-5);
        	\draw[->] (m-2-4) -- (m-3-5);
        	\draw[->] (m-1-5) -- (m-2-6);
        	\draw[->] (m-3-5) -- (m-2-6);
        	\draw[->] (m-2-6) -- (m-3-7);
        	\draw[red] (a) -- (m-3-1);
        	\draw[red] (a) -- (m-3-3);
        	\draw[red] (a) -- (m-2-2);
        	\draw[red] (b) -- (m-2-2);
        	\draw[red] (b) -- (m-2-4);
        	\draw[red] (b) -- (m-1-3);
        	\draw[red] (b) -- (m-3-3);
        	\draw[red] (c) -- (m-1-3);
        	\draw[red] (c) -- (m-1-5);
        	\draw[red] (c) -- (m-2-4);
        	\draw[red] (d) -- (m-3-3);
        	\draw[red] (d) -- (m-3-5);
        	\draw[red] (d) -- (m-2-4);
        	\draw[red] (e) -- (m-2-4);
        	\draw[red] (e) -- (m-2-6);
        	\draw[red] (e) -- (m-1-5);
        	\draw[red] (e) -- (m-3-5);
        	\draw[red] (f) -- (m-3-5);
        	\draw[red] (f) -- (m-3-7);
        	\draw[red] (f) -- (m-2-6);
        \end{tikzpicture}
    \end{adjustbox}
	\caption{The facet defining inequalities of the type cone~$\typeCone(\gvectorFan[n])$ correspond to the flips described in \cref{prop:extremalExchangeablePairsAsso}. See \cref{subsec:typeConeCA,sec:clusterCategories} for a representation theoretic \mbox{interpretation}.}
	\label{fig:labelFacetDefiningInequalititesAsso}
\end{figure}
\end{example}

@@

\begin{example}
Consider the cluster fans illustrated in \cref{fig:clusterFans}.
The type cone of the left fan of \cref{fig:clusterFans} lives in~$\R^9$ and has a linearity space of dimension~$3$.
It has $6$ facet defining inequalities (given below), which correspond to the mesh mutations of \cref{thm:extremalExchangeablePairsCA} as illustrated in \cref{fig:labelFacetDefiningInequalititesCA}\,(top).

@@

\part{Grothendieck group and relations between $\b{g}$-vectors}

@@

\subsection{Statement of the theorem}\label{subsect::Statement cluster cats}
We need to introduce some notations before stating the main result of this section.  Let~$\Lambda\eqdef \End{\cat}(T)$, and let~$F$ be the functor
\[
F = \cat(T,-): \cat \xrightarrow{} \MOD \Lambda.
\]

@@

\begin{definition}\label{defi::grothendieck-group}
 \begin{enumerate}
  \item Let~$\Ksp(\cat)$ be the \defn{split Grothendieck group} of~$\cat$, that is, the free abelian group generated by symbols~$[X]$, where~$[X]$ denotes the isomorphism class of~$X$ in~$\cat$, modulo the following relations: for any objects~$Y$ and~$Z$, we let~$[Y\oplus Z] = [Y] + [Z]$.

@@

\begin{definition}\label{defi::bilinear form}
 For any two objects~$X$ and~$Y$ of~$\cat$, define
 \[
  \langle X, Y \rangle \eqdef \dim_{\field} \Hom{\Lambda}(FX,FY).
 \]
 This defines a bilinear form
 \[
  \langle -,-\rangle : \Ksp(\cat) \times \Ksp(\cat) \xrightarrow{} \Z.
 \]
\end{definition}

>>>>

% formating the part command
\makeatletter
\def\part{\@startsection{part}{1}%
\z@{.7\linespacing\@plus\linespacing}{.8\linespacing}%
{\LARGE\sffamily\centering}}
%\@addtoreset{section}{part}
\makeatother
\renewcommand{\thepart}{\Roman{part}}
%\renewcommand{\thesection}{\arabic{part}.\arabic{section}}

@@

\vspace{-.5cm}

@@

\section*{Introduction}

@@

The associahedron appears as a fundamental structure in several mathematical theories, such as moduli spaces and topology~\cite{Stasheff, Keller-AinfinityAlgebras}, operads and rewriting theory~\cite{}, combinatorial Hopf algebras~\cite{LodayRonco, ChatelPilaud, Pilaud-brickAlgebra}, diagonal harmonics~\cite{}, mathematical physics~\cite{ArkaniHamedBaiHeYan}, etc.
Another striking illustration of the ubiquity of the associahedron lies in the theory of cluster algebras introduced by S.~Fomin and A.~Zelevinsky in~\cite{FominZelevinsky-ClusterAlgebrasI} with motivation coming from total positivity and canonical bases.
Cluster algebras are commutative rings generated by variables obtained from an initial seed by the discrete dynamical process of seed mutations.
Finite type cluster algebras, whose mutation graph is finite, were classified in~\cite{FominZelevinsky-ClusterAlgebrasII} using the Cartan-Killing classification for crystallographic root systems.
Finite type cluster complexes were geometrically realized first by F.~Chapoton, S.~Fomin and A.~Zelevinsky~\cite{ChapotonFominZelevinsky} based on the $\b{d}$-vector fans~\cite{FominZelevinsky-YSystems, FominZelevinsky-ClusterAlgebrasII} of the bipartite initial seed, then by~C.~Hohlweg, C.~Lange and H.~Thomas~\cite{HohlwegLangeThomas} based on the Cambrian fans~\cite{Reading-CambrianLattices, ReadingSpeyer} of acyclic initial seeds, and finally by C.~Hohlweg, V.~Pilaud and S.~Stella~\cite{HohlwegPilaudStella} based on the $\b{g}$-vector fans~\cite{FominZelevinsky-ClusterAlgebrasIV} with respect to an arbitrary initial seed.
The resulting polytopes are known as generalized associahedra, and the classical associahedra corresponds to the cluster algebras of type~$A$.

This paper focuses on a surprising construction of the associahedron that recently appeared in mathematical physics.
Motivated by the prediction of the behavior of scattering particules, N.~Arkani-Hamed, Y.~Bai, S.~He, and G.~Yan recently described in~\cite[Sect.~3.2]{ArkaniHamedBaiHeYan} the kinetic associahedron.
It is a class of polytopal realizations of the classical associahedron obtained as sections of a high-dimensional positive orthant with well-chosen affine subspaces.
This construction provides a large degree of freedom in the choice of the parameters defining these affine subspaces and produces all polytopes whose normal fan is affinely equivalent to that of J.-L.~Loday's associahedron~\cite{Loday} (see \cref{subsec:typeConeAsso}).
These realizations were then extended by V.~Bazier-Matte, G.~Douville, K.~Mousavand, H.~Thomas and E.~Yildirim~\cite{BazierMatteDouvilleMousavandThomasYildirim} in the context of finite type cluster algebras using tools from representation theory of quivers.
More precisely, they fix a finite type cluster algebra~$\c{A}$ and consider the real euclidean space~$\R^\c{V}$ indexed by the set~$\c{V}$ of cluster variables of~$\c{A}$.
Starting from an acyclic initial seed~$\seed_\circ$, they consider a certain set~$\c{M}$ of mesh mutations and compute the corresponding linear dependences between the $\b{g}$-vectors, that can be interpreted as linear spaces in~$\R^\c{V}$.
Finally, they perturb these mesh linear spaces by a collection~$\b{\ell} \in \R^{\c{M}}$ of positive parameters and intersect the resulting perturbed mesh affine spaces with the positive orthant of~$\R^\c{V}$ (see \cref{subsec:typeConeCA}).
They show that the resulting polytope is always a generalized associahedron, and that its normal fan is the $\b{g}$-vector fan of~$\c{A}$ with respect to the initial acyclic seed~$\seed_\circ$.
A by-product of their construction is that the space of all polytopal realizations of the $\b{g}$-vector fan of~$\c{A}$ with respect to the initial acyclic seed~$\seed_\circ$ is a simplicial cone.

In this paper, we revisit, extend and explore further this construction using an reversed approach.
Given a complete simplicial fan~$\Fan$, we consider the space~$\typeCone(\Fan)$ of all its polytopal realizations, called type cone of~$\Fan$ in~\cite{McMullen-typeCone}.
This type cone is known to be a polyhedral cone defined by a collection of inequalities corresponding to the linear dependences among the rays of~$\Fan$ contained in pairs of adjacent maximal cones of~$\Fan$ (see \cref{def:typeCone}).
Our approach is based on an elementary but powerful observation: if the type cone~$\typeCone(\Fan)$ is a simplicial cone, then all polytopal realizations of~$\Fan$ can be described as sections of a high dimensional positive orthant with a collection of affine subspaces corresponding to the facets of~$\typeCone(\Fan)$ (see \cref{coro:simplicialTypeCone}).
To prove that the type cone~$\typeCone(\Fan)$ is simplicial, we just need to identify which pairs of adjacent maximal cones of~$\Fan$ correspond to the facets of~$\typeCone(\Fan)$ and to show that the corresponding linear dependences among their rays positively span the linear dependence among the rays of any pair of adjacent maximal cones of~$\Fan$.
When applied to the $\b{g}$-vector fans of cluster algebras (see \cref{subsec:typeConeCA}), this reversed approach enables us to efficiently revisit and extend the construction of~\cite{BazierMatteDouvilleMousavandThomasYildirim}.
This new perspective has several advantages:
\item Our proof applies to any initial seed, regardless of whether it is acyclic or not (see \cref{exm:typeConeCA,fig:clusterFans,fig:labelFacetDefiningInequalititesCA} for examples in types~$A_3$ and~$C_3$~cyclic). In contrast, the proof of~\cite{BazierMatteDouvilleMousavandThomasYildirim} only treat acyclic initial seeds.
\item Our proof is uniform for all finite type cluster algebras, regardless of whether it is simply-laced or not (see \cref{exm:typeConeCA} for an example in type~$C_3$). In contrast, even with the acyclicity restriction, the result of~\cite{BazierMatteDouvilleMousavandThomasYildirim} is first proved in simply-laced cases ($A$, $D$ and~$E$) using representation theory of quivers, and then argued by a folding argument. Although not presented in detail, this argument is subtle and technical.
\item Our proof directly applies to any positive real valued parameters, regardless of whether they are rational or not. In contrast, the proof of~\cite{BazierMatteDouvilleMousavandThomasYildirim} naturally applies to rational parameters and then requires to apply an approximation argument.
These advantages all follow from one essential feature of our reversed approach.
Namely, it enables to completely separate the algebraic and the geometric aspects of the problem:
\item On the algebraic side, we study the relations between the $\b{g}$-vectors of a finite type cluster algebra and we show that the mesh mutations minimally generate these relations.
For this, we use representation theory in the setting of a~$2$-Calabi--Yau triangulated categories, and we adapt M.~Auslander's proof that the relations in the Grothendieck group of an Artin algebra are generated by the ones given by almost-split sequences precisely when the algebra is of finite representation type~\cite{Auslander1984} (see \cref{}).
\item On the geometric side, we observe that our algebraic result precisely states that all type cones of $\b{g}$-vector fans of cluster algebras are simplicial and we just use elementary manipulations to show that a fan whose type cone is simplicial is affinely equivalent to the normal fan of sections of a high dimensional positive orthant with well-chosen affine subspaces (see \cref{}).
These two aspects are interesting for their own right and might be useful in different communities: the geometric side enables to describe all polytopal realizations of a fan with simplicial type cone (which is sometimes easier to check than to find an explicit realization), while the algebraic side provides fundamental new results on the $\b{g}$-vectors and the Grothendieck groups of cluster algebras.
We have therefore voluntarily split our presentation into two clearly marked parts that can be read independently.
The geometric side, which uses elementary techniques, is presented in \cref{part:geometry} together with all its polytopal realization consequences.
The algebraic side, which is our core contribution, is presented in \cref{part:algebra}.

@@

% other generalizations would deserve the same study.

@@

\part{Type cones of $\b{g}$-vector fans}
\label{part:geometry}

@@

\begin{definition}
\label{def:typeCone}
The \defn{type cone} of~$\Fan$ is the cone~$\typeCone(\Fan)$ of all $\Fan$-admissible height vectors~$\b{h}$:
\begin{align*}
\typeCone(\Fan) & \eqdef \set{\b{h} \in \R^N}{\Fan \text{ is the normal fan of } P_\b{h}} \\
& = \Bigset{\b{h} \in \R^N}{\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0 \; \begin{array}{l} \text{for any adjacent maximal} \\ \text{cones~$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$} \end{array}}.
\end{align*}
\end{definition}

@@

Although we will only treat with simplicial fans, for completeness, note that \cref{prop:characterizationPolytopalFan} and \cref{def:typeCone} can be easily adapted to non-simplicial fans. To describe the type cone of an arbitrary complete fan, it suffices to consider any simplicial refinement, and to set some of the strict inequalities of the definition of the type cone to equality.

\begin{proposition}
\label{prop:non-simplicial}
Let $\Fan'$ be a complete fan that coarsens the essential complete simplicial fan~$\Fan$.
The \defn{type cone}~$\typeCone(\Fan')$ of all $\Fan'$-admissible height vectors~$\b{h}$ is
\begin{align*}
\typeCone(\Fan) 
%  & \eqdef \set{\b{h} \in \R^N}{\Fan \text{ is the normal fan of } P_\b{h}} \\
& = \left\{\b{h} \in \R^N\;\middle|\;
\begin{array}{ll}
\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} = 0
& \begin{array}{l} \text{for any adjacent maximal cones} \\ \text{$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ \textbf{belonging}}\\\text{to the same maximal cone of $\Fan'$; and}\end{array}
\\\\
\sum_{\ray[s] \in \rays \cup \rays'} \coefficient[{\ray[s]}][\rays][\rays'] \, \b{h}_{\ray[s]} > 0
& \begin{array}{l} \text{for any adjacent maximal cones} \\ \text{$\R_{\ge0}\rays$ and~$\R_{\ge0}\rays'$ of~$\Fan$ \textbf{not belonging}}\\\text{to the same maximal cone of $\Fan'$}\end{array}
\end{array}
\right\}.
\end{align*}
\end{proposition}
\arnau{should add proof or citation}

@@

\begin{example}
\label{ex:typecone}
Consider the $2$-dimensional fan~$\Fan$ depicted in \cref{fig:typecone} (left) (we will encounter this fan again in \cref{exm:typeConeAsso}). It has five rays $v_1,\dots,v_5$ and five maximal cones $a,\dots, e$. Its type cone lies in~$\R^5$, but has a $2$-dimensional linearity space. It suffices to study $\b{K}\ctypeCone(\Fan)$ which is an essential simplicial cone in~$\R^3$. One can further reduce the dimension by intersecting with a transversal hyperplane, to get a $2$-dimensional representation. This is the triangle that is depicted in \cref{fig:typecone} (right). The five pairs of adjacent maximal cones give rise to five defining inequalities for $\ctypeCone(\Fan)$, although only three are non-redundant. If $H_{ij}$ is the hyperplane supporting the inequality given by the two adjacent maximal cones~$i$ and~$j$, then only $H_{ab}$, $H_{cd}$ and $H_{bc}$ support facets of the type cone. The relative interior of the three facets of~$\ctypeCone(\Fan)$ are the type cones of three coarsenings of~$\Fan$, obtained by merging certain adjacent maximal cones (those that are extremal). One of the rays generating~$\b{K}\ctypeCone(\Fan)$ corresponds to a $2$-dimensional simplex, and its normal fan is obtained by merging two pairs of extremal adjacent maximal cones. The two remaining rays lie not only in the hyperplanes~$H_{ij}$ defining the facets it belongs to, but also one extra such hyperplane. Its normal fan has only two cones, one arising as a merging of three cones and one arising as a merging of two cones. These fans are not essential, they have a one-dimensional linearity. Hence, the corresponding polytopes are not full-dimensional, they are of codimension~$1$.

\begin{figure}[htpb]
	\capstart
	\centerline{{\includegraphics[width=.3\linewidth]{typeconeFan}}\hspace{1.5cm}\includegraphics[width=.4\linewidth]{typecone}}
	\caption{A complete $2$-dimensional fan~$\Fan$ with five rays $v_1,\dots,v_5$ and five maximal cones $a,\dots, e$. We depict the intersection of $\b{K}\ctypeCone(F)$ with a hyperplane to get a $2$-dimensional slice. On top of each face of $\ctypeCone(F)$, and example of polytope with the corresponding normal fan is depicted.}
	\label{fig:typecone}
\end{figure}
\end{example}

@@

\begin{figure}[h]
	\capstart
    \begin{adjustbox}{center}
        \begin{tikzpicture}
        	\matrix (m) [matrix of math nodes, row sep=.35cm, column sep=.2cm, nodes={anchor=center, align=center, inner sep=0pt}]{
        		& \;\includegraphics[scale=.7]{diagonalA2} & \node (b) {\red \circled{B}}; & \;\includegraphics[scale=.7]{diagonalA4} & \\
        		\;\includegraphics[scale=.7]{diagonalA1} & \node (a) {\red \circled{A}}; & \;\includegraphics[scale=.7]{diagonalA3} & \node (c) {\red \circled{C}}; & \;\includegraphics[scale=.7]{diagonalA5} \\};
        	\draw[->] (m-2-1) -- (m-1-2);
        	\draw[->] (m-1-2) -- (m-2-3);
        	\draw[->] (m-2-3) -- (m-1-4);
        	\draw[->] (m-1-4) -- (m-2-5);
        	\draw[red] (a) -- (m-2-1);
        	\draw[red] (a) -- (m-2-3);
        	\draw[red] (a) -- (m-1-2);
        	\draw[red] (b) -- (m-1-2);
        	\draw[red] (b) -- (m-1-4);
        	\draw[red] (b) -- (m-2-3);
        	\draw[red] (c) -- (m-2-3);
        	\draw[red] (c) -- (m-2-5);
        	\draw[red] (c) -- (m-1-4);
        \end{tikzpicture}
        \quad
        \begin{tikzpicture}
        	\matrix (m) [matrix of math nodes, row sep=.35cm, column sep=.2cm, nodes={anchor=center, align=center, inner sep=0pt}]{
        		&& \;\includegraphics[scale=.7]{diagonalB3} & \node (c) {\red \circled{C}}; & \;\includegraphics[scale=.7]{diagonalB7} && \\
        		& \;\includegraphics[scale=.7]{diagonalB2} & \node (b) {\red \circled{B}}; & \;\includegraphics[scale=.7]{diagonalB5} & \node (e) {\red \circled{E}}; & \;\includegraphics[scale=.7]{diagonalB8} & \\
        		\;\includegraphics[scale=.7]{diagonalB1} & \node (a) {\red \circled{A}}; & \;\includegraphics[scale=.7]{diagonalB4} & \node (d) {\red \circled{D}}; & \;\includegraphics[scale=.7]{diagonalB6} & \node (f) {\red \circled{F}}; & \;\includegraphics[scale=.7]{diagonalB9} \\};
        	\draw[->] (m-3-1) -- (m-2-2);
        	\draw[->] (m-2-2) -- (m-1-3);
        	\draw[->] (m-2-2) -- (m-3-3);
        	\draw[->] (m-1-3) -- (m-2-4);
        	\draw[->] (m-3-3) -- (m-2-4);
        	\draw[->] (m-2-4) -- (m-1-5);
        	\draw[->] (m-2-4) -- (m-3-5);
        	\draw[->] (m-1-5) -- (m-2-6);
        	\draw[->] (m-3-5) -- (m-2-6);
        	\draw[->] (m-2-6) -- (m-3-7);
        	\draw[red] (a) -- (m-3-1);
        	\draw[red] (a) -- (m-3-3);
        	\draw[red] (a) -- (m-2-2);
        	\draw[red] (b) -- (m-2-2);
        	\draw[red] (b) -- (m-2-4);
        	\draw[red] (b) -- (m-1-3);
        	\draw[red] (b) -- (m-3-3);
        	\draw[red] (c) -- (m-1-3);
        	\draw[red] (c) -- (m-1-5);
        	\draw[red] (c) -- (m-2-4);
        	\draw[red] (d) -- (m-3-3);
        	\draw[red] (d) -- (m-3-5);
        	\draw[red] (d) -- (m-2-4);
        	\draw[red] (e) -- (m-2-4);
        	\draw[red] (e) -- (m-2-6);
        	\draw[red] (e) -- (m-1-5);
        	\draw[red] (e) -- (m-3-5);
        	\draw[red] (f) -- (m-3-5);
        	\draw[red] (f) -- (m-3-7);
        	\draw[red] (f) -- (m-2-6);
        \end{tikzpicture}
    \end{adjustbox}
	\caption{The facet defining inequalities of the type cone~$\typeCone(\gvectorFan[n])$ correspond to the flips described in \cref{prop:extremalExchangeablePairsAsso}. See \cref{subsec:typeConeCA,sec:clusterCategories} for a representation theoretic \mbox{interpretation}.}
	\label{fig:labelFacetDefiningInequalititesAsso}
\end{figure}
\end{example}

@@

\begin{example}
\label{exm:typeConeCA}
Consider the cluster fans illustrated in \cref{fig:clusterFans}.
The type cone of the left fan of \cref{fig:clusterFans} lives in~$\R^9$ and has a linearity space of dimension~$3$.
It has $6$ facet defining inequalities (given below), which correspond to the mesh mutations of \cref{thm:extremalExchangeablePairsCA} as illustrated in \cref{fig:labelFacetDefiningInequalititesCA}\,(top).

@@

\part{Grothendieck group and relations between $\b{g}$-vectors}
\label{part:algebra}

@@

\subsection{Statement of the theorem}
\label{subsect::Statement cluster cats}

We need to introduce some notations before stating the main result of this section.  Let~$\Lambda\eqdef \End{\cat}(T)$, and let~$F$ be the functor
\[
F = \cat(T,-): \cat \xrightarrow{} \MOD \Lambda.
\]

@@

\begin{definition}
\label{defi::grothendieck-group}
 \begin{enumerate}
  \item Let~$\Ksp(\cat)$ be the \defn{split Grothendieck group} of~$\cat$, that is, the free abelian group generated by symbols~$[X]$, where~$[X]$ denotes the isomorphism class of~$X$ in~$\cat$, modulo the following relations: for any objects~$Y$ and~$Z$, we let~$[Y\oplus Z] = [Y] + [Z]$.

@@

\begin{definition}
\label{defi::bilinear form}
 For any two objects~$X$ and~$Y$ of~$\cat$, define
 \[
  \langle X, Y \rangle \eqdef \dim_{\field} \Hom{\Lambda}(FX,FY).
 \]
 This defines a bilinear form
 \[
  \langle -,-\rangle : \Ksp(\cat) \times \Ksp(\cat) \xrightarrow{} \Z.
 \]
\end{definition}
